Poisson approximate likelihood compared to the particle filter

### 2024-09-12 ms.Rnw
another two rounds

<<estimated-mle_stocks_init, echo=FALSE, message=FALSE, warning=FALSE>>=
# Can be changed if more rounds needed
est_mle_vec_sbh <- sbh_el_05[which.max(sbh_el_05$logLik),]
@

@@

\begin{table}[htbp]
    \caption{Maximum likelihood estimation of parameters for the OvOv model of WWR by PAL (taken from WWR), and by iterated particle filtering implemented using {\rm \texttt{mif2}} in {\rm \texttt{pomp}} \citep{pomppackagepaper}. The initial distribution parameter \(\lambda_0 = (S_{10}, I_{10}, R_{10}, S_{20}, I_{20}, R_{20}, S_{30}, I_{30}, R_{30})\) are assumed to be fixed in \cite{wwr}. The iterated particle filtering estimates were obtained by using 3 rounds of refinement, with 100 filtering iterations, 50,000 particles, and 36 replicates in each round with the top 12 best fits in terms of likelihood chosen to be the starting value for the next round.}
    \label{tab:mlebywwr}
\begin{center}    
    \begin{tabular}{lccc}  % Left-aligned for the first column, centered for the other two columns
      \toprule
      Parameter      & PAL      & iterated filtering & iterated filtering (stocks)\\
      \midrule
      \(\beta_1\)    & 11.48     & $\Sexpr{round(est_mle_vec["beta1"],2)}$ & $\Sexpr{round(est_mle_vec_sbh["beta1"],2)}$\\
      \(\beta_2\)    & 0.25      & $\Sexpr{round(est_mle_vec["beta2"],2)}$ & $\Sexpr{round(est_mle_vec_sbh["beta2"],2)}$\\
      \(\beta_3\)    & 0.35      & $\Sexpr{round(est_mle_vec["beta3"],2)}$ & $\Sexpr{round(est_mle_vec_sbh["beta3"],2)}$\\
      \(\phi\)       & 0.14      & $\Sexpr{round(est_mle_vec["phi"],2)}$   & $\Sexpr{round(est_mle_vec_sbh["phi"],2)}$\\
      \(\rho\)       & 0.16      & $\Sexpr{round(est_mle_vec["beta11"],2)}$ & $\Sexpr{round(est_mle_vec_sbh["beta11"],2)}$\\
      \(\sigma^2_q\) & 0.021     & $\Sexpr{round(est_mle_vec["sigma_q"],3)}$ & $\Sexpr{round(est_mle_vec_sbh["sigma_q"],3)}$\\
      \(\sigma_\xi\) & 66.89     & $\Sexpr{round(est_mle_vec["sigma_xi"],2)}$ & $\Sexpr{round(est_mle_vec["sigma_xi"],2)}$\\
      \(S_{10}\)     & 3876549   & $\Sexpr{round(pop*est_mle_vec["S10"],0)}$  & $5094840$\\
      \(I_{10}\)     & 30351     & $\Sexpr{round(pop*est_mle_vec["I10"],0)}$  & $2871$\\
      \(R_{10}\)     & 1315221   & $\Sexpr{round(pop*est_mle_vec["R10"],0)}$  & $124410$\\
      \(S_{20}\)     & 57139612  & $\Sexpr{round(pop*est_mle_vec["S20"],0)}$  & $57385600$\\
      \(I_{20}\)     & 871       & $\Sexpr{round(pop*est_mle_vec["I20"],0)}$  & $639$\\
      \(R_{20}\)     & 302852    & $\Sexpr{round(pop*est_mle_vec["R20"],0)}$  & $57072$\\
      \(S_{30}\)     & 19573727  & $\Sexpr{round(pop*est_mle_vec["S30"],0)}$  & $19697600$\\
      \(I_{30}\)     & 2550      & $\Sexpr{round(pop*est_mle_vec["I30"],0)}$  & $174$\\
      \(R_{30}\)     & 131092    & $\Sexpr{round(pop*est_mle_vec["R30"],0)}$  & $9578$\\
      \midrule
      \textbf{AIC}  & 13778.08  & $\Sexpr{round(2*((length(params_wwr) - length(params_fixed)) - est_mle_vec["logLik"]),2)}$ & $\Sexpr{round(2*((length(params_wwr) - length(params_fixed)) - est_mle_vec_sbh["logLik"]),2)}$\\
      \bottomrule
    \end{tabular}
\end{center}
\end{table}

>>>>

<<maximization6-Stocks_rinit, echo=F, warning=F, message=F>>=
set.seed(12345)
sd_06 = (2/3)^5
starting_values_sbh_06 <- choose_n_fits(sbh_el_05, top_n_fits, ncores)

bake(file = paste0(out_dir, "mif2_maximization/SBH_init/round_06/mifs_global_sbh_06.rds"),{ 
  plan(multicore, workers = ncores)
  foreach(i = 1:ncores, .options.future = list(seed = 652643293)) %dofuture% {
    pomp::mif2(
      mifs_global_sbh_05[[1]],
      rw.sd = rw_sd(
        beta1 = 0.01*sd_06, beta2 = 0.01*sd_06, beta3 = 0.01*sd_06,
        beta11 = 0.01*sd_06, phi = 0.01*sd_06,
        sigma_q = 0.01*sd_06, sigma_xi = 0.01*sd_06
      ),
      params = starting_values_sbh_06[[i]]
    ) 
  }
}) -> mifs_global_sbh_06

bake(file = paste0(out_dir,"mif2_maximization/SBH_init/round_06/sbh_el_06.rds"),{
  eval_logLik_pomp(mifs_global_sbh_06, ncores, NP_MIF, seed = 4248930)
}) ->  sbh_el_06
@

<<maximization7-Stocks_rinit, echo=F, warning=F, message=F>>=
set.seed(12345)
sd_07 = (2/3)^6
starting_values_sbh_07 <- choose_n_fits(sbh_el_06, top_n_fits, ncores)

bake(file = paste0(out_dir, "mif2_maximization/SBH_init/round_07/mifs_global_sbh_07.rds"),{ 
  plan(multicore, workers = ncores)
  foreach(i = 1:ncores, .options.future = list(seed = 652643293)) %dofuture% {
    pomp::mif2(
      mifs_global_sbh_06[[1]],
      rw.sd = rw_sd(
        beta1 = 0.01*sd_07, beta2 = 0.01*sd_07, beta3 = 0.01*sd_07,
        beta11 = 0.01*sd_07, phi = 0.01*sd_07,
        sigma_q = 0.01*sd_07, sigma_xi = 0.01*sd_07
      ),
      params = starting_values_sbh_07[[i]]
    ) 
  }
}) -> mifs_global_sbh_07

bake(file = paste0(out_dir,"mif2_maximization/SBH_init/round_07/sbh_el_07.rds"),{
  eval_logLik_pomp(mifs_global_sbh_07, ncores, NP_MIF, seed = 4248930)
}) ->  sbh_el_07
@

<<estimated-mle_stocks_init, echo=FALSE, message=FALSE, warning=FALSE>>=
# Can be changed if more rounds needed
est_mle_vec_sbh <- sbh_el_07[which.max(sbh_el_07$logLik),]
@

@@

\begin{table}[htbp]
    \caption{Maximum likelihood estimation of parameters for the OvOv model of WWR by PAL (taken from WWR), and by iterated particle filtering implemented using {\rm \texttt{mif2}} in {\rm \texttt{pomp}} \citep{pomppackagepaper}. The initial distribution parameter \(\lambda_0 = (S_{10}, I_{10}, R_{10}, S_{20}, I_{20}, R_{20}, S_{30}, I_{30}, R_{30})\) are assumed to be fixed in \cite{wwr}. The iterated particle filtering estimates were obtained by using 3 rounds of refinement, with 100 filtering iterations, 50,000 particles, and 36 replicates in each round with the top 12 best fits in terms of likelihood chosen to be the starting value for the next round.}
    \label{tab:mlebywwr}
\begin{center}    
    \begin{tabular}{lccc}  % Left-aligned for the first column, centered for the other two columns
      \toprule
      Parameter      & PAL      & iterated filtering & iterated filtering (stocks)\\
      \midrule
      \(\beta_1\)    & 11.48     & $\Sexpr{round(est_mle_vec["beta1"],2)}$ & $\Sexpr{round(est_mle_vec_sbh["beta1"],2)}$\\
      \(\beta_2\)    & 0.25      & $\Sexpr{round(est_mle_vec["beta2"],2)}$ & $\Sexpr{round(est_mle_vec_sbh["beta2"],2)}$\\
      \(\beta_3\)    & 0.35      & $\Sexpr{round(est_mle_vec["beta3"],2)}$ & $\Sexpr{round(est_mle_vec_sbh["beta3"],2)}$\\
      \(\phi\)       & 0.14      & $\Sexpr{round(est_mle_vec["phi"],2)}$   & $\Sexpr{round(est_mle_vec_sbh["phi"],2)}$\\
      \(\rho\)       & 0.16      & $\Sexpr{round(est_mle_vec["beta11"],2)}$ & $\Sexpr{round(est_mle_vec_sbh["beta11"],2)}$\\
      \(\sigma^2_q\) & 0.021     & $\Sexpr{round(est_mle_vec["sigma_q"],3)}$ & $\Sexpr{round(est_mle_vec_sbh["sigma_q"],3)}$\\
      \(\sigma_\xi\) & 66.89     & $\Sexpr{round(est_mle_vec["sigma_xi"],2)}$ & $\Sexpr{round(est_mle_vec_sbh["sigma_xi"],2)}$\\
      \(S_{10}\)     & 3876549   & $\Sexpr{round(pop*est_mle_vec["S10"],0)}$  & $5094840$\\
      \(I_{10}\)     & 30351     & $\Sexpr{round(pop*est_mle_vec["I10"],0)}$  & $2871$\\
      \(R_{10}\)     & 1315221   & $\Sexpr{round(pop*est_mle_vec["R10"],0)}$  & $124410$\\
      \(S_{20}\)     & 57139612  & $\Sexpr{round(pop*est_mle_vec["S20"],0)}$  & $57385600$\\
      \(I_{20}\)     & 871       & $\Sexpr{round(pop*est_mle_vec["I20"],0)}$  & $639$\\
      \(R_{20}\)     & 302852    & $\Sexpr{round(pop*est_mle_vec["R20"],0)}$  & $57072$\\
      \(S_{30}\)     & 19573727  & $\Sexpr{round(pop*est_mle_vec["S30"],0)}$  & $19697600$\\
      \(I_{30}\)     & 2550      & $\Sexpr{round(pop*est_mle_vec["I30"],0)}$  & $174$\\
      \(R_{30}\)     & 131092    & $\Sexpr{round(pop*est_mle_vec["R30"],0)}$  & $9578$\\
      \midrule
      \textbf{AIC}  & 13778.08  & $\Sexpr{round(2*((length(params_wwr) - length(params_fixed)) - est_mle_vec["logLik"]),2)}$ & $\Sexpr{round(2*((length(params_wwr) - length(params_fixed)) - est_mle_vec_sbh["logLik"]),2)}$\\
      \bottomrule
    \end{tabular}
\end{center}
\end{table}
